---
title: "Class05: Data Viz with ggplot2"
author: "Julissa Gonzalez (PID:A18495188)"
format: pdf
toc: true
---

## Background

There are many graphic systems in R for making plots and figures
These include so-called *"base R" graphics* like the `plot()` function and on packages like**ggplot2**.

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:

```{r}
head(cars)
```
```{r}
plot(cars)
```
Before I can use ggplot I need to install it on my computer.
To do this we can use the function `install.packages("ggplot2")`
>**N.B** We never run `instal.packages()` in our quarto doc we run it once only in our R console as it would re-install the package every time we render out quarto report
once installed we need to load up the package into our R brain:

```{r}
library(ggplot2)

```

The main function in the **ggplot2** package is called `ggplot()
```{r}
ggplot(cars)
```



Every ggplot has at least 3 layers:
the **data** (a data.frame of the stuff we want to plot)
the **aes** thetics(how the data maps to the plot),
the **geom** layer (how you want the plot drawn, e.g. points, lines, etc.)
```{r}
ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point()
```
## Add some custom features
Lets add a trend line that shows the relationship between speed and distance.

```{r}
ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point() +
  geom_line()

  
```
```{r}
ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point() +
  geom_smooth()+
  theme_bw()+
  labs(titles="Stoping Dist of old cars",
       x="Speed (MPH)",
       y= " Distance (ft)")
```
Q. Can you make the `geom_smooth()`function produce a linear
```{r}
ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point() +
  labs(titles="Stoping Dist of old cars",
       x="Speed (MPH)",
       y= " Distance (ft)")+
   geom_smooth(method="lm", se=FALSE)+
  theme_bw()
```
## Gene expression Figure
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
first plot attempt
```{r}
ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()
```

```{r}
ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()+
  scale_colour_manual(values=c("purple","grey","orange"))+  labs(x="No Drug ",
      y="Drug ",
      title="Gene Expresion upon GLP-1 inhibitor treatment")


```
## Expression Chnages
```{r}

#File location 
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (i.e.rows) are in this dataset?

```{r}
nrow(gapminder)

```

> Q. How many "country" are in this dataset?

```{r}
length(table(gapminder$country))

```

```{r}
length(unique(gapminder$country))
```
let's make our fist plot of the entire dataset:

plot of "gdpPercap" vs. "lifeExp" colored by "continent"

```{r}
p<- ggplot(gapminder)+
  aes(gdpPercap,lifeExp,col=continent )+
  geom_point(alpha=0.3)
p
```
I can add more layers to 'p'

```{r}
p+
  facet_wrap(~continent)

```
Make a plot for 1977 and 2007
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
library(dplyr)
gapminder_both<-gapminder%>% filter(year==1977|year==2007)

gapminder_2007<-gapminder%>% filter(year==2007)

```
```{r}
ggplot(gapminder_2007)+
  aes(gdpPercap,lifeExp)+
  geom_point()
```


```{r}
ggplot(gapminder_both)+
  geom_point(aes(x=gdpPercap, y=lifeExp, col=continent, size=pop),alpha=0.7)+
  facet_wrap(~year)
```
```{r}
ggplot(gapminder)+
  aes(lifeExp,col=continent)+
  geom_histogram()
  

```

```{r}
ggplot(gapminder)+
  aes(lifeExp,fill=continent)+
  geom_histogram()
```

Q. Make a histogram of lifeExp facted by continent 

```{r}
ggplot(gapminder)+
  aes(lifeExp)+
  geom_histogram()+
  facet_wrap(~continent)
```
