---
title: "Class06: R functions"
author: "Julissa Gonzalez (A18495188)"
format: pdf
toc: true
---

## Background

All functions in R have at least 3 things:

- a *name* (we pick that and use it to call the function)
- input *arguments* ( one or more comma separated inputs that go inside the bracket when we call the function),
- the *body* ( the lines of R code that do the work of the function).

## A first function

Here we will create a function to add some numbers. Let's call it `add()`

Input arguments can be either "required" or "optional. The later have
fall-back defualt values that will be used if the user does not specify
them.

```{r}
add<- function(...){
  return(sum(... ))
}

```

can wew use our new function:

```{r}
add(10,100)
```

```{r}
add(10)
```

> **Q1A.** Your first version of the function should add two input
> numbers together. For example, add(4,7) should return 11

```{r}
add(4, 7)
```

> **Q1B.** For you second version, adapt your first function so it can
> take a single input vector or two inputs as before. For example,
> add(4, 7) and add( c(4,7,10) )

```{r}
add(4, 7)
add( c(4,7,10) )
```

> **Q1C.** Finally, on your own (outside of class) create a third
> version of your function that can add any number of inputs that the
> user provides. For example, add(1, 2, 3, -4) should return

```{r}
add(1, 2, 3, -4)
```

```{r}
ans<- add(1,2,3)
ans
```

We can explicitly set a **return** value output from a function (rather
that the default last time)

```{r}
add<- function(x,y=0,z=0){
  return(sum(x,y,z))
  cat("Is it break time yet?\n")
}

add(10,100)
```

can we use our new function:

## A `generate_dna()` function

a useful function here is the "base R" `sample()` function:

```{r}

sample(1:5,size=60, replace=TRUE)
```

We can use this to make a random nucleotide sequwnce if we draw from
"A","C", "G", and "T"...

```{r}
sample(x=c("A","C","G","T"), size=10, replace=TRUE)
```

> **Q2.** Write a function generate_dna() that returns a random DNA
> sequence of a length specified by the user. Your first version should
> return a multi-element vector of single character nucleotides. For
> example generate_dna(6) might return "A", "T", "T", "G", "A", "C". \[1
> pt\]

```{r}
generate_dna <- function(len=10) {
  sample(x=c("A","C","G","T"), size=len, replace=TRUE)
}
```

```{r}
generate_dna()
generate_dna(100)
```

> **Q2b**:Your second version should *optionally* be able to return
> either a multi-element vector of single character nucleotides (as
> before) or a **single character string** (not a vector of single
> letters but a singe vector of multiple letters). For example
> "AAGGCTG". \[1 pt\]

```{r}
generate_dna <- function(len=10,single.element=TRUE) {
  ans<-sample(c("A","C","G","T"), size=len, replace=TRUE)
  #cat("Hello...")
  if(single.element){
    #cat("Is it me your looking for...")
   ans<- paste(ans, collapse= "")
  }
  
  ##Format as FASTA with an ID line
  cat(paste(">len",len,"\n", sep=""))
  cat(ans)
  cat("\n")
  
  ##
  ##
  return(ans)
}
```

Functions that could be useful here are `paste()`, `if()`, `cat()`and
`return()`

```{r}
generate_dna(44)
```

```{r}
generate_dna(single.element = TRUE)
```

```{r}
generate_dna(single.element=FALSE)
```

```{r}
paste(c("A","C","G"),collapse="---")
```

> **Q2C**.Finally, create a final version of your function that prints out a FASTA format sequence with an id line indicating the sequence length. For example:

```         
>len9
CGAAGGCTG
```

```{r}
cat("hello \n there")
```

```{r}
generate_dna()
```

## Write a `generate protien()` function

> **Q3.** Write a function generate_protein() that returns a random
> peptide/protein sequence of a length specified by the user. For
> example generate_protein(6) might return "WQRTAG".

```{r}
generate_protien<-function(len=9){
  aa<-c("A","R","N","D","C","E","Q","G","H","I","L","K","K","M","F","P","S","T","W","Y","V")
  ans<-sample(x=aa, size=len, replace=TRUE)
  paste(ans,collapse="")
  }
```

```{r}
generate_protien(6)
```

## Write a \`Generate random protien sequences of length 6 to 13

> **Q4.**Adapt and use your generate_protein() function to generate a
> series of random protein sequences ranging from 6 to 13 amino acids in
> length (one sequence per length). Take advantage of the base R
> function for() or sapply() so that you do not have to call
> generate_protein() eight times by hand.

```{r}
for(l in 6:13){
  cat(">", l, "\n", sep="")
  cat(generate_protien(l), "\n")
}
```

```{r}
generate_protien(6)
generate_protien(7)
generate_protien(8)
generate_protien(9)

```

## Are Our peptides

> **Q5**:We will define a peptide as “unique in nature” (in the specific sense used in this lab) if no match exists with 100% coverage AND 100% identity to a sequence already in nr. Note that this is a conservative definition — a short exact match may still exist as a sub-string of a longer protein, and will typically be reported by BLASTp with 100% identity but less than 100% coverage of the subject (which we are not using here).

| Length | Ide | Cov | Unique |
|--------|-----|-----|--------|
| 6      | 100 | 100 | N      |
| 7      | 100 | 100 | N      |
| 8      | 100 | 100 | N      |
| 9      | 100 | 88  | Y      |
| 10     | 100 | 80  | Y      |
| 11     | 88  | 82  | Y      |
| 12     | 100 | 67  | Y      |
| 13     | 88  | 69  | Y      |

## Connecting your findings to immunology (MHC class II and T-cell activation) 

>**Q6.** In 3–6 sentences total and using your Q5 data and the reasoning from Q5b, what do you think
this minimum length is and why might it be a bad design choice for the immune system to present
very short peptides? [3 pt]

Based Q5 results the minimum peptide length is around 9 amino acids, since shorter peptides 6-8 were not unique. Peptides of length 9 and above began to show uniqueness. This means that shorter sequences are more likely to occur by chance and match existing proteins. Short peptides would be a poor  choice for the immune system because they may not be specific enough to distinguish between self and non-self proteins. This could increase the risk of false recognition or failure to detect pathogens. Longer peptides provide greater specificity, improving accurate immune responses

