---
title: "HW6"
author: "Julissa Gonzalez"
date: "2026-04-18"
output: pdf_document
---


## Q6
### Inputs
- `pdb_id`: PDB ID or file name
- `chain`: chain to analyze (default = `"A"`)
- `elety`: atom type to analyze (default = `"CA"`)

### What the function does and how to use it
This function reads in protein PDB data, trims it to the specified chain and atom type, extracts the B-factors, and outputs a plot for the specified protein.  
Use it by calling `plot_bfactor("4AKE")` or another PDB ID.


### Output
The function outputs a B-factor plot for the specified protein and returns a numeric vector of B-factor values.

```{r}
library(bio3d)

plot_bfactor <- function(pdb_id, chain = "A", elety = "CA") {
  pdb <- read.pdb(pdb_id)
  pdb_trim <- trim.pdb(pdb, chain = chain, elety = elety)
  bvals <- pdb_trim$atom$b
  plotb3(bvals, sse = pdb_trim, typ = "l", ylab = "Bfactor")
  return(bvals)
}

plot_bfactor("4AKE")
```

### Example output

```{r}
s1.b <- plot_bfactor("4AKE")
s2.b <- plot_bfactor("1AKE")
s3.b <- plot_bfactor("1E4Y")
```




